---
title: 'ANLY 560: Data Viz and EDA'
author: "Merykokeb Belay"
date: "2/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(tseries)
library(forecast)
library(astsa) 
library(xts)
library(fpp2)
library(fma)
library(lubridate)
library(tidyverse)
library(TSstudio)
library(quantmod)
library(tidyquant)
library(plotly)
library(ggplot2)
```



```{r}
#import data from COVID tracking project; 
county_covid<-read.csv('covid-county-by-race.csv')
county_covid
```

# DATA VIZ AND EDA
```{r}
#plot counties with the top-5 covid-19 deaths 
county_covid_plot<-county_covid%>%
  arrange(desc(deathsPer100k))%>%
  select(state, countyName, deathsPer100k)%>%
  mutate(location=paste(countyName,',',state))%>%
  slice(1:5)%>%
  ggplot(aes(deathsPer100k, reorder(location, deathsPer100k)))+geom_bar(stat='identity', width = 0.6, fill='#fc8d62', alpha=0.6)+theme_minimal()+labs(y="", x='Deaths per 100k', title='U.S. COVID-19 Deaths per 100k: Top Five Counties', subtitle = 'The graph below depicts the top five U.S.counties with the highest COVID-19 \ndeaths. At the time of data collection, Galax city, Virginia had the highest\nnumber of deaths recorded.', caption='Data source: The COVID-19 Tracking Project')+theme(plot.title = element_text(size = 15, margin = margin(b = 10), hjust=0), plot.subtitle = element_text(size = 10, color = "darkslategrey", margin = margin(b = 25), hjust=0), plot.caption = element_text(size=6))
county_covid_plot
#ggsave('county_covid_plot.jpeg')
```

The plot above depicts counties with the highest covid-19 deaths, as recroded by the COVID-19 tracking project. These counties are found in Virginia, South Dakota, Georgia, and Texas.

Let's look at the state of hospitals in these states using data provided by Health and Human Services. 


```{r}
df<-read_csv('COVID-19_Reported_Patient_Impact_and_Hospital_Capacity_by_State_Timeseries.csv')
df
```

```{r}
Sys.setenv("plotly_username"="mbelay")
Sys.setenv("plotly_api_key"="ETtjOXrmELd3uq5bhkfY")

```


```{r}
plotly_df<-df%>%
  mutate(critical_staffing_total= critical_staffing_shortage_today_yes+critical_staffing_shortage_today_no+critical_staffing_shortage_today_not_reported)%>%
  select(state, date, critical_staffing_shortage_today_yes, critical_staffing_shortage_today_no, critical_staffing_shortage_today_not_reported, critical_staffing_total)%>%
  mutate(perc_critical_staffing_shortage_yes=(critical_staffing_shortage_today_yes/critical_staffing_total)*100)%>%
  filter(state=='VA'| state=='TX'| state=='SD'| state=='GA')%>%
  arrange(date)

fig1<-plot_ly(plotly_df%>%filter(state=='VA'), x=~date, y=~perc_critical_staffing_shortage_yes, type='scatter', mode='lines', name = 'VA')#%>%layout(title='Hospital Critical Staff Shortage: January 2020 - February 2022', yaxis=list(title='Percent of hospital with reported shorages'), xaxis=list(title=''))
fig2<-plot_ly(plotly_df%>%filter(state=='GA'), x=~date, y=~perc_critical_staffing_shortage_yes, type='scatter', mode='lines', name='GA')#%>%layout(title='Hospital Critical Staff Shortage: January 2020 - February 2022', yaxis=list(title='Percent of hospital with reported shorages'), xaxis=list(title=''))
fig3<-plot_ly(plotly_df%>%filter(state=='SD'), x=~date, y=~perc_critical_staffing_shortage_yes, type='scatter', mode='lines', name='SD')#%>%layout(title='Hospital Critical Staff Shortage: January 2020 - February 2022', yaxis=list(title='Percent of hospital with reported shorages'), xaxis=list(title=''))
fig4<-plot_ly(plotly_df%>%filter(state=='TX'), x=~date, y=~perc_critical_staffing_shortage_yes, type='scatter', mode='lines', name='TX')
fig<-subplot(fig1, fig2, fig3, fig4, nrows=4, shareX = TRUE, titleY =FALSE)#%>%layout(title='Hospital Critical Staff Shortage: January 2020 - February 2022',xaxis=list(title=''))

m <- list(
  l = 100,
  r = 25,
  b = 25,
  t = 125
  #pad = 4
)
fig <- fig %>% layout(title=list(text='Hospital Critical Staff Shortage: January 2020 - February 2022'),
                      
                      #width = 500, height = 500,
                      margin = m)

  
# your axis title
fig <- fig %>% layout(annotations = list(
                list(x = -0.15 , y = 0.5, text = "Percent of hospitals reporting shortages",
                     
                     textangle = 270,
                     showarrow = F, xref='paper', yref='paper', size=48)
                
                ))
fig
api_create(fig, filename = 'us_hospital_shortages')
```



```{r}
plotly_df<-df%>%
  select(state, date, percent_of_inpatients_with_covid)%>%
  mutate(perc_of_inpatients_with_covid= percent_of_inpatients_with_covid*100)%>%
  filter(state=='VA'| state=='TX'| state=='SD'| state=='GA')%>%
  arrange(date)

fig<-plot_ly(plotly_df, x=~date, y=~percent_of_inpatients_with_covid, group_by=~state,color = ~state, type='scatter', mode='lines')
  
fig
```



```{r}
#plot inpatients with covid in texas

tx_covid<-df%>%
  select(state, date, percent_of_inpatients_with_covid)%>%
  mutate(perc_of_inpatients_with_covid= (percent_of_inpatients_with_covid)*100)%>%
  filter(state=='TX')%>%
  arrange(date)%>%
  ggplot(aes(x=date, y=perc_of_inpatients_with_covid))+geom_line(color='#fc8d62')+theme_minimal()+labs(y='Percent of Inpatients with COVID-19', x='', title='Share of Inpatients with COVID-19 in Texas: Jan 2020 - Feb 2022', subtitle = "The line plot below depicts the share of inpatients with COVID-19 across hospitals in Texas' \nhealthcare system. Although there is an initial trend to highlight the immediate rise in inpatient cases \nfollowing the start of the pandemic, it tends to be less pronounced over time")+theme(plot.title = element_text(size = 15, margin = margin(b = 10), hjust=0), plot.subtitle = element_text(size = 10, color = "darkslategrey", margin = margin(b = 25), hjust=0), plot.caption = element_text(size=6))+ylim(c(0,40))
#ggsave('tx_covid_plot.jpeg')
tx_covid

```



```{r}
#plot inpatients with covid in texas

tx_covid<-df%>%
  select(state, date, deaths_covid_coverage)%>%
  filter(state=='VA'| state=='TX'| state=='SD'| state=='GA')%>%
  arrange(date)%>%
  ggplot(aes(x=date, y=deaths_covid_coverage, group=state)) + 
  geom_line(aes(color=state))+theme_minimal()+labs(y='Number of hospitals reporting COVID-19 deaths', x='', title='Hospitals reporting COVID-19 deaths: Jan 2020 - Feb 2022', subtitle = "The line plot below depicts the number of hospital facilities that reported covid deaths\n from January 2020 to February 2022. Compared to Georgia, Sout Dakota, and Virigina,\n hospitals in Texas were reporting far more deaths.")+theme(plot.title = element_text(size = 15, margin = margin(b = 10), hjust=0), plot.subtitle = element_text(size = 10, color = "darkslategrey", margin = margin(b = 25), hjust=0), plot.caption = element_text(size=6))
ggsave('tx_covid_plot.jpeg')
tx_covid

```


### Examine uninsured rates across these states 

```{r}

uninsured<-read_csv('uninsured.csv')
uninsured
```

```{r}
uninsured_plot<-uninsured%>%
  pivot_longer(cols = c('2008', '2009', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018', '2019'))%>%
  mutate(value=value*100)%>%
  ggplot(aes(x=name, y=value, group=Location)) +
  geom_line(aes(color=Location))+theme_minimal()+labs(y='Percent of Uninsured Population', x='', title='Health Insurance Coverage of Total Population: 2008 - 2019', subtitle = "The line plot below depicts the share of the total population that is uninsured in Georgai, \nTexas, Virginia, and South Dakota.")+theme(plot.title = element_text(size = 15, margin = margin(b = 10), hjust=0), plot.subtitle = element_text(size = 10, color = "darkslategrey", margin = margin(b = 25), hjust=0), plot.caption = element_text(size=6))+ylim(c(0, 25))
#ggsave('uninsured_plot.jpeg')
```





```{r}
#create a univariate time series dataset
tx_covid_inpatient<-df%>%
  select(state, date, percent_of_inpatients_with_covid)%>%
  filter(state=='TX')%>%
  mutate(perc_of_inpatients_with_covid= percent_of_inpatients_with_covid*100)%>%
  select(date, p=perc_of_inpatients_with_covid)
tx_covid_inpatient<-tx_covid_inpatient[order(tx_covid_inpatient$date),]
td=tx_covid_inpatient$p
str(td)
#turn into a time series format
td.ts<-ts(td, start = c(2020, 1, 1),frequency = 7)
#plot series
plot(td.ts)
#decompose series
plot(decompose(td.ts))
#plot autocorrelation 
plot(acf(diff(td.ts)))
str(td.ts)
#display lag plot
gglagplot(td.ts,lags=12, do.lines = FALSE)
#ggsave('lags.jpeg')

```


```{r}
#select multiplicative decomposition 
plot(decompose(td.ts, "multiplicative"))

```


```{r}
#plot acf and pacf
acf(td.ts)
pacf(td.ts)

```

```{r}
#stationarity
adf.test(td.ts)
acf(diff(diff(td.ts)))
pacf(diff(td.ts), main='Partial Autocorrelation Function Plot (First Difference)')
adf.test(diff(td.ts))
```


```{r}
#moving average smoothing 
autoplot(td.ts)+autolayer(ma(td.ts, order=7), series = 'MA-7')+autolayer(ma(td.ts, order=29), series='MA-29')+autolayer(ma(td.ts, order=89), series = 'MA-89')+autolayer(ma(td.ts, order=179), series='MA-179')+theme_minimal()+ylab('Share of Hospitals Reporting Shortages')+labs(title = 'Texas Reported Hospital Critical Staff Shortages')+xlab('')

```



#ARMA/ARIMA

```{r}

model_smry<-data.frame()
#begin for loop to try different combinations
for (p in 1:3){
  for (q in 1:4){
    for (d in 1:2){
      fit<-arima(td.ts, order=c(p,d,q))
      acc<-accuracy(fit)
      acc_ext<-data.frame(p,d,q, AIC=AIC(fit), BIC=BIC(fit))
      model_smry<-rbind(model_smry, acc_ext)
    }
  }
}
model_smry%>% 
  filter(d==1)%>%
  mutate(summ=p+d+q)%>%
  filter(summ<=8)%>%
  arrange(AIC)



```



```{r}

model_train<-Arima(td.ts, order=c(1,1,1), include.drift = TRUE)
summary(model_train)
```

```{r}

autoplot(forecast(model_train,100), PI=F, seris='dfd')+geom_line()+ylab('Hospital Critical Staff Shortages (Texas)')+ xlab('')

```


```{r}

#mean
f1 <- meanf(td.ts, h=100)
#naive
f2 <- naive(td.ts, h=100)

#Drift
f3 <- rwf(td.ts, h=100, drift = TRUE)


#seasonal naive
f4<-snaive(td.ts, h=100)


```

```{R}
autoplot(td.ts)+autolayer(pred$mean, PI=FALSE, series='ARIMA(1,1,1)') + autolayer(meanf(td.ts, h=100),
series="Mean", PI=FALSE) + autolayer(naive(td.ts, h=100),
series="Na誰ve", PI=FALSE) + autolayer(rwf(td.ts, drift=TRUE, h=100),
series="Drift", PI=FALSE) + autolayer(snaive(td.ts, h=100), PI=FALSE, series='Seasonal Na誰ve')+
ggtitle("Texas Hospital Critical Staff Shortages\n(daily ending 8 Feb 2022)") + xlab("") + ylab("Hospital Critical Staff Shortages (Texas)") + guides(colour=guide_legend(title="Forecast"))

```

```{r}

pred=forecast(model_train,100)
autoplot(pred$mean)

```


### Seasonal ARIMA 

```{r}

#seasonal differencing 
td.ts %>%ggtsdisplay()
td.ts %>%diff()%>%ggtsdisplay()
td.ts %>% diff(lag=7) %>%ggtsdisplay()
td.ts %>% diff(lag=7) %>% diff()%>%ggtsdisplay()
acf(diff(td.ts))

```

```{r}
#fit model
sarima(td.ts,4,1,3,1,1,1,7)
sarima(td.ts,4,1,1,1,1,1,7)



```


```{R}


model_smry<-data.frame()
#begin for loop to try different combinations
for (p in 1:6){
  for (q in 1:6){
    for (d in 1:2){
      fit<-arima(td.ts, order = c(p,d,q),seasonal = list(order=c(1,1,1), period=7) )
      acc<-accuracy(fit)
      acc_ext<-data.frame(p,d,q, AIC=AIC(fit), BIC=BIC(fit))
      model_smry<-rbind(model_smry, acc_ext)
    }
  }
}
model_smry%>% 
  filter(d==1)%>%
  mutate(summ=p+d+q)%>%
  filter(summ<=8)%>%
  arrange(AIC)




```


```{r}

fit<-arima(td.ts, order = c(4,1,1),seasonal = list(order=c(1,1,1), period=7) )
res.arima1<-fit$res
acf(res.arima1)
pacf(res.arima1)
sarima(res.arima1, 9,0,0)
```


```{r}

autoplot(forecast(fit,100), PI=T, seris='dfd')+geom_line()+ylab('Hospital Critical Staff Shortages (Texas)')+ xlab('')

```


```{r}

#mean
f1<-meanf(td.ts, h=100)
#naive
f2 <- naive(td.ts, h=100)

#Drift
f3 <- rwf(td.ts, h=100, drift = TRUE)


#seasonal naive
f4<-snaive(td.ts, h=100)


```

```{r}

pred=forecast(fit,100)
autoplot(pred$mean)
summary(pred)

```


```{R}
autoplot(td.ts)+autolayer(pred$mean, PI=FALSE, series='SARIMA(4,1,1,1,1,1)7') + autolayer(meanf(td.ts, h=100),
series="Mean", PI=FALSE) + autolayer(naive(td.ts, h=100),
series="Na誰ve", PI=FALSE) + autolayer(rwf(td.ts, drift=TRUE, h=100),
series="Drift", PI=FALSE) + autolayer(snaive(td.ts, h=100), PI=FALSE, series='Seasonal Na誰ve')+
ggtitle("Texas Hospital Critical Staff Shortages\n(daily ending 8 Feb 2022)") + xlab("") + ylab("Hospital Critical Staff Shortages (Texas)") + guides(colour=guide_legend(title="Forecast"))

```




# Financial Time Series

```{r, message=FALSE, warning=FALSE}
#install.packages("grDevices")
#library(grDevices)
library(fGarch) 
library(tseries)
library(astsa) 
library(xts)
library(fpp)
library(astsa)
library(ggplot2)
library(tidyverse)

```


```{r, message=FALSE, warning=FALSE}

library(quantmod)
getSymbols("PFE", from="2022-02-01", src="yahoo")
head(`PFE`)
autoplot(PFE$PFE.Adjusted, main='Pfizer Adjusted Price Feb-March 2022', xlab='')
```



```{r}

#candlestick plot of prices 
chartSeries(`PFE`, type="candlesticks", theme="white")

#plot returns 
head(btc.close<- Ad(`PFE`),na.action = na.omit)
returns = diff(log(btc.close))
chartSeries(returns, theme="white")

```



```{r}

#acf of return 
acf(returns, na.action = na.pass, lag.max = 100)
acf(abs(returns), na.action = na.pass, lag.max = 100)
acf((returns)^2, na.action = na.pass, lag.max = 100)


#pacf of returns
pacf(returns, na.action = na.pass, lag.max = 100)
pacf(abs(returns), na.action = na.pass, lag.max = 100)
pacf((returns)^2, na.action = na.pass, lag.max = 100)


```

```{r}
library(fGarch)
x<-garchFit(~garch(3,0), na.remove(returns))
y<-garchFit(~garch(4,0), na.remove(returns))


```

```{r}
autoplot(as.ts(x@residuals))

```


```{r}


model_smry<-data.frame()
#begin for loop to try different combinations
for (p in 1:2){
  for (q in 0:4){
    for (d in 0:1){
      fit<-arima(returns, order = c(p,d,q))
      acc<-accuracy(fit)
      acc_ext<-data.frame(p,d,q, AIC=AIC(fit), BIC=BIC(fit))
      model_smry<-rbind(model_smry, acc_ext)
    }
  }
}
model_smry



```

```{r}

fit2<-arima(returns, order=c(2,0,0))
sarima(na.omit(returns), 2,0,0)


```

```{r}

res.arima1=fit2$res
squared.res.arima1=res.arima1^2


#par(mfcol=c(3,1))
plot(squared.res.arima1,main='Squared Residuals')
acf.squared1=acf(squared.res.arima1,main='ACF Squared Residuals',lag.max=100, na.action = na.omit)
pacf.squared1=pacf(squared.res.arima1,main='PACF Squared Residuals',lag.max=100, na.action = na.omit)




```

```{r}

acf((returns)^2, na.action = na.pass, lag.max=100)
pacf((returns)^2, na.action = na.pass, lag.max=100)
acf(abs(returns), na.action = na.pass, lag.max=100)
pacf(abs(returns), na.action = na.pass, lag.max=100)


```

```{r}

library(FinTS)
byd.archTest <- ArchTest(returns, lags = 1, demean = TRUE)
byd.archTest
```


There are no ARCH effects. 

```{r}
x<-garchFit(~garch(3,0), na.remove(returns))
#summary(garch(na.remove(returns),order=c(0,3),trace=F))
x@fit$coef

```
$\delta=-0.0027936332$
$var(y_{t}|y_{t-1})=\sigma^2=0.0001373019+0.0000000100y^2_{t-1}+0.0411269692y^2_{t-2}+0.4902650183y^2_{t-3}$


### Seasonal cross validation

```{r}
autoplot(td.ts)
```



```{r}
k <- 48 # minimum data length for fitting a model (4 seasonal lags)
n <- length(td.ts) #144
n-k #96
mae1 <- matrix(NA,n-k,7)
mae2 <- matrix(NA,n-k,7)


st <- tsp(td.ts)[1]+(k-2)/7 #ending point 
#tsp=time series period

a0 <- window(td.ts, st=tsp(td.ts)[1],end=st )
a0
```

```{r}
k <- 75 # minimum data length for fitting a model 
n <- length(td.ts)
n-k 
for(i in 1:(n-k))
{
  xtrain <- td.ts[1:(k-1)+i]
  xtest <- td.ts[k+i]
  
  fit <- Arima(xtrain, order=c(4,1,1), seasonal = list(order=c(1,1,1), period=7))
  fcast1 <- forecast(fit, h=1)
  
  fit2 <- Arima(xtrain, order=c(4,1,2), seasonal = list(order=c(1,1,1), period=7))
  fcast2 <- forecast(fit2, h=1)
  
  
  err1 = c(err1, abs(fcast1$mean-xtest))
  err2 = c(err2, abs(fcast2$mean-xtest))
  
}

mean(err1) 

```


```{r}

farima1 <- function(td.ts, h){forecast(Arima(td.ts, order=c(4,1,1), seasonal = list(order=c(1,1,1), period=7)), h=h)}
e <- tsCV(td.ts, farima1, h=7)
length(e)
```


```{R}


# Compute the MSE values and remove missing values
df<-data.frame(e=c(e), time=c(time(e)))
mse <- colMeans(df, na.rm = TRUE)

# Plot the MSE values against the forecast horizon
data.frame(h = 0:7, MSE = mse) %>%
  ggplot(aes(x = h, y = MSE)) + geom_point()+labs(title = 'Seven-step ahead')



```











